\section{$\RTR$ Credential Chain Discovery}
\label{section-chain-discovery}

In this section we discuss an algorithm for authorization with risk in a distributed
environment. As in $\RT$ chain discovery, \cite{Li:2003-02}, a central issue is that credentials
need not be stored locally, but may need to be retrieved from remote machines. Furthermore,
modulation of tolerable risk is intended to be a central feature of $\RTR$; online authorization
need not be risk-optimal, but risk should be kept below a certain threshold. This allows
tolerable levels of risk to be specified and enforced. This flexibility will be implemented by
viewing authorization decisions as a reachability analysis in a \emph{credential graph}, and by
defining chain discovery as a partial graph reconstruction algorithm parameterized by a
\emph{maximum risk threshold}.

\subsection{Credential Graphs}

We begin by defining an interpretation of credential sets $\creds$ as a credential graph. More
precisely, sets of credentials are interpreted as a weighted multigraph, where nodes are role
expressions, edges are credentials, and weights are risks. Authorization is implemented by
determining reachability, via \emph{risk weighted paths}, where the aggregation of edge risk
along the path is the risk of authorization. Reachability is predicated on simple paths, since
traversing cycles can only increase risk, and any path with a cycle would otherwise generate an
infinite number of risk weighted paths. Allowing the latter would preclude a constructive
definition of credential graphs, as will become clear below.
\begin{definition}[Risk weighted paths]
  Let $\graph{} = (\nodes{}, \wtedges{})$ be a weighted multigraph with nodes $f \in \nodes{}$
  and edges $\wtedge{f_1}{f_2}{\risk} \in \wtedges{}$ weighted by elements $\risk$ of a given
  risk ordering. The pair:
$$((f_1,\ldots,f_n), \risk_1 \oplus \cdots \oplus \risk_{n-1})$$ is a
\emph{risk weighted path in $\graph{}$} iff for all $i \in [1..n-1]$, there exists
$\wtedge{f_i}{f_{i+1}}{\risk_i} \in \wtedges{}$. A weighted path $((f_1,\ldots,f_n), \risk)$ is
\emph{simple} iff no node is repeated in $(f_1,\ldots,f_n)$. We write $\wtpath{f}{f'}{\risk}$,
pronounced ``$f'$ is reachable from $f$ with risk $\risk$'', iff $((f,\ldots,f'), \risk)$ is a
simple risk weighted path. We write $\wtpath{f}{f'}{\risk} \in \graph{}$ iff
$\wtpath{f}{f'}{\risk}$ holds given $\graph{}$.
\end{definition}

Now, the definition of credential graphs is founded on the definition of risk weighted paths,
since edges derived from linked and intersection credentials are supported by them.
\begin{definition}[Credential graph]
  \label{def-credentialgraph}
  Given $\creds$, its \emph{credential graph} is a weighted multigraph $\graph{\creds} =
  (\nodes{\creds}, \wtedges{\creds})$, where:
$$
\nodes{\creds} = \bigcup\limits_{\cred{A.r}{e}{\risk}} \setdefn{A.r,e}
$$
And $\wtedges{\creds}$ is the least set of risk-weighted edges satisfying the following closure
properties:
\begin{enumerate}
\item If $\cred{A.r}{e}{\risk} \in \creds$ then $\wtedge{e}{A.r}{\risk} \in \wtedges{\creds}$.
\item If $B.r_2, A.r_1.r_2 \in \nodes{\creds}$ and $\wtpath{B}{A.r_1}{\risk}$ is simple, then
  $\wtedge{B.r_2}{A.r_1.r_2}{\risk} \in \wtedges{\creds}$.
\item If $D,f_1\cap \cdots \cap f_n \in \nodes{\creds}$ and for each $i \in [1..n]$ there exists
  a simple path $\wtpath{D}{f_i}{\risk_i}$, then $\wtedge{D}{f_1 \cap \cdots \cap f_n}{\risk}
  \in \wtedges{\creds}$, where $\risk = \risk_1 \oplus \cdots \oplus \risk_n$.
\end{enumerate}
\end{definition}

The definition of credential graphs can be made constructive by iterating closure over an
initial initial edge set $\wtedges{\creds}^0$:
$$
\wtedges{\creds}^0 = \setdefn{\wtedge{A.r}{e}{\risk} \mid 
\cred{A.r}{e}{\risk} \in \creds}
$$ 
In rules (2) and (3), the paths predicating membership in $\wtedges{\creds}$ are called
\emph{support paths}, and the edges are called \emph{derived}. On each iteration, add a new
weighted edge according to closure rule (2) or (3). Since $\creds$ is finite, and support paths
must be simple, the process will reach a fixpoint in a finite number of iterations; this
fixpoint is $\wtedges{\creds}$.

We observe that the characterization of credential sets $\creds$ is sound and complete with
respect to the set theoretic semantics given in the previous section. These results will form a
bridge with the semantics of $\RTR$ for establishing correctness of credential chain discovery.
The statement of soundness reflects the fact that while risk assessments of credential sets
express minimum risk bounds of role membership, the credential graph does not preclude
reachability via paths of higher risk.
\begin{theorem}[Soundness]
  For all $B,A.r$, if $\wtpath{B}{A.r}{\risk} \in \graph{\creds}$, then $(B,\risk') \in
  \credsmean{\creds}(A.r)$ with $\risk' \po \risk$.
\end{theorem}
The statement of completeness reflects that any assessed risk is the weight of some related path
in the graph:
\begin{theorem}[Completeness]
  For all $A.r$, if $(B,\risk) \in \credsmean{\creds}(A.r)$, then $\wtpath{B}{A.r}{\risk} \in
  \graph{\creds}$.
\end{theorem}

\subsection{Backward Chain Discovery Algorithm}

The credential chain discovery algorithm is a modification of the backward search algorithm in
\cite{Li:2003-02}. For a particular risk lattice $\mathcal{K}$ and risk aggregration operation
$\oplus$, the algorithm takes a collection of credentials $\mathcal{C}$, the name of a governing
role, and a maximum risk $\risk_M \in \mathcal{K}$. The algorithm returns a risk assessment for
the governing role where each entity in the assessment is associated with a risk $\risk \po
\risk_M$.

The algorithm maintains a set of credential graph nodes $N$ and edges $E$, adding nodes and
edges to these sets as it works. The algorithm also maintains a queue of nodes $Q$ that are
waiting to be processed. When this queue is exhausted the algorithm terminates. Associated with
each node in $N$ is a set of aggregated risk records (ARRs) of the form $(n_d, n_s, \risk_A)$.
The driving node $n_d$ contains the name of the node that caused the ARR to be initially
created. The support node $n_s$ is used when handling intersection roles. The aggregated risk
$\risk_A$ reflects the risk accumulated from the point where the ARR is first created. In what
follows we use $N(e)$ to denote the set of ARRs in node $e$.

We define a canonical representation for sets of ARRs $\mathcal{A}$ analogously to risk
assessments. $\mathcal{A}$ is in canonical form if there does not exist $(n_d, n_s, \risk_1),
(n_d, n_s, \risk_2) \in \mathcal{A}$ such that $\risk_1 \po \risk_2$. In addition we extend the
definition of $\oplus$ to include sets of ARRs
$$
\mathcal{A} \oplus \risk' = \{(n_d, n_s, \risk \oplus \risk')\,|\, (n_d, n_s, \risk) \in
\mathcal{A}\}
$$

It is convenient to define an operation $\mathit{update}(\mathcal{A}, \risk, e_2)$ that
describes how a set of ARRs $\mathcal{A}$ are copied to a node $e_2$ over an edge in the
credential graph with risk $\risk$. The $\mathit{update}$ function first adds a node for $e_2$
to the graph if one does not already exist. It then computes
$$
A' = \mathit{canonical}(\mathit{bound}(\mathcal{A} \oplus \risk) \cup N(e_2))
$$
where $\mathit{canonical}$ is a function that returns a set of ARRs that is the canonical form
of the set given as its argument, and where $\mathit{bound}$ is a function that removes all ARRs
$(n_d, n_s, \risk)$ from the given set where $\risk \po \risk_M$ is false. If $N(e_2) \ne A'$
then $\mathit{update}$ sets $N(e_2) = A'$ and queues $e_2$ for exploration.

The algorithm processes each type of node, except for entity nodes, as follows
\begin{description}
\item[Role $A.r$] Look up all edges $\cred{A.r}{e}{\risk}$. For each such edge,
  $\mathit{update}(N(A.r), \risk, e)$.
\item[Linked role $A.s.t$] Look up all edges $\cred{A.s.t}{e}{\risk}$. For each such edge,
  $\mathit{update}(N(A.r.s), \risk\ e)$. In addition $\mathit{update}((A.s.t, A.s, \bot), \bot,
  A.s)$.
\item[Intersection role $f_1 \cap \cdots \cap f_n$] Look up all edges $\cred{f_1 \cap \cdots
    \cap f_n}{e}{\risk}$. For each such edge, $\mathit{update}(N(f_1 \cap \cdots \cap f_n),
  \risk, e)$. In addition, for each $f_i$ participating in the intersection
  $\mathit{update}((f_1 \cap \cdots \cap f_n, f_i, \bot), \bot, f_i)$.
\end{description}

These rules describe how ARRs propagate through the credential graph. The algorithm starts by
creating and queuing a node for the governing role $g$ and assigning to that node the ARR $(g,
g, \bot)$. As this ARR moves through the graph it accumulates risk information about the various
paths to $g$. If two paths converge, the node at the junction of the paths will get re-explored
if the second set of incoming ARRs cause accumulated risks at that node to change.

Maintaining the canonical form of the ARR sets in each node keeps the algorithm from diverging
when there are cycles in the credential graph. When the edge $\cred{e_1}{e_2}{\risk}$ that
closes the cycle is processed, the $\mathit{update}$ operation will not modify $N(e_2)$ since
all ARRs in $N(e_1) \oplus \risk$ will have risk values greater than or equal to those in
$N(e_2)$. Thus there will be no change to the canonical form of $N(e_2)$ and $e_2$ will not be
re-explored.

If at any time during the search the total aggregated risk $\risk$ in an ARR is such that $\risk
\po \risk_M$ is false, that ARR no longer propagates and the search in that direction is
abandoned. Thus the algorithm does not bother exploring portions of the credential graph that
have too much aggregated risk from the governing role.

Entity nodes must be processed in a special way so that the derived edges they imply are
properly handled. The operation $\mathit{processEntity}$ loops over all ARRs in the entity node
$E$ and for each ARR $(n_d, n_s, \risk)$ considers one of three cases
\begin{description}
\item[$n_d$ is a plain role $A.r$] In this case, $A.r$ must be the governing role and $E$ can
  reach that node with an aggregated risk $\risk$. The algorithm records this discovery.
\item[$n_d$ is a linked role $A.s.t$] In this case $E$ can reach $A.s$ with risk $\risk$. The
  algorithm adds the edge $\cred{A.s.t}{E.t}{\risk}$ to the edge set and queues $A.s.t$ for
  re-exploration.
\item[$n_d$ is an intersection role $f_1 \cap \cdots \cap f_n$] In this case $E$ can reach $n_s$
  with risk $\risk$. The algorithm associates this information with the pair $(n_d, E)$. If all
  roles mentioned in $n_d$ have been associated with $(n_d, E)$ in this way, the algorithm adds
  the edge $\cred{f_1 \cap \cdots \cap f_n}{E}{\risk}$ to the edge set and queues $f_1 \cap
  \cdots \cap f_n$ for re-exploration.
\end{description}

To illustrate the action of the algorithm consider the sum-of-risks example in Section
\ref{sec:sum-of-risks}. In that example $\mathit{Ed}$ is in the \role{Store}{buyer} role both
with risk 8 and through the \role{Personnel}{manager} role with risk 9. If $\risk_M = 8$, the
algorithm returns a risk assessment for the \role{Store}{buyer} role of $\{(\mathit{Ed}, 8)\}$.
The path with risk 9 is not fully explored because the total aggregated risk on that path
exceeds $\risk_M$. The algorithm processes this example as follows.
\begin{enumerate}

\item The node for \role{Store}{buyer} is created an initialized with an ARR of $(S.b, S.b,
  \bot)$.

\item When \role{Store}{buyer} is processed, a node is created for \role{Acme}{purchaser} $\cap$
  \role{Acme}{employee} and the ARR in \role{Store}{buyer} is copied to the new node and updated
  to $(S.b, S.b, 1)$.

\item When \role{Acme}{purchaser} $\cap$ \role{Acme}{employee} is processed, there are no edges
  yet defining it. However, nodes for \role{Acme}{purchaser} and \role{Acme}{employee} are
  created and initialized with $(A.p \cap A.e, A.p, \bot)$ and $(A.p \cap A.e, A.e, \bot)$
  respectively.

\item When \role{Acme}{purchaser} is processed, nodes for \role{Personnel}{manager} and
  $\mathit{Ed}$ are created. The ARRs in \role{Acme}{purchaser} are copied to these new nodes
  and updated resulting in $(A.p \cap A.e, A.p, 2) \in N(P.m)$ and $(A.p \cap A.e, A.p, 4) \in
  N(E)$.

\item When \role{Acme}{employee} is processed, the ARR $(A.p \cap A.e, A.e, 3)$ is added to
  $N(E)$

\item When \role{Personnel}{manager} is processed, the ARR $(A.p \cap A.e, A.p, 5)$ is computed
  from $N(P.m) \oplus 3$. However, this ARR is removed by the canonicalization process and thus
  is never added to $N(E)$.

\item When $\mathit{Ed}$ is processed $\mathit{processEntity}$ loops over the ARRs contained in
  $N(E)$. It finds that the intersection is satisfied and thus creates the edge $\cred{A.p \cap
    A.e}{E}{7}$. It then queues node $A.p \cap A.e$ for re-exploration.

\item When \role{Acme}{purchaser} $\cap$ \role{Acme}{employee} is processed again the algorithm
  copies the ARR $(S.b, S.b, \bot)$ to $E$ and updates it to $(S.b, S.b, 8)$. Since this changes
  $N(E)$ that node is queued for re-exploration. However, \role{Acme}{purchaser} and
  \role{Acme}{employee} are not queued since their ARR sets are not modified.

\item When $\mathit{Ed}$ is processed again, the discovery is made that $E$ is in $S.b$ with
  risk 8. The other ARRs in $N(E)$ are also processed causing the node for $A.p \cap A.e$ to be
  re-explored. However, none of the canonical ARR sets are modified after that and thus the
  queue drains and the algorithm terminates.
\end{enumerate}

\begin{theorem}
  If $\mbox{\texttt{checkmem}}(E, A.r, \risk)$ returns true, then there exists $(E, \risk') \in
  \mathcal{S}_\mathcal{C}(A.r)$ such that $\risk' \po \risk$.
\end{theorem}

\begin{theorem}
  For all $(E, \risk) \in \mathcal{S}_\mathcal{C}(A.r)$, $\mbox{\texttt{checkmem}}(E, A.r,
  \risk)$ returns true.
\end{theorem}

