%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% FILE         : main.tex
% AUTHOR       : (C) Copyright 2007 by Peter C. Chapin
% LAST REVISED : 2007-03-27
% SUBJECT      : Presentation given at 5AIASS
%
% This is a presentation on RT^R.
%
% Send comments or bug reports to:
%
%       Peter C. Chapin
%       University of Vermont
%       Burlington, VT 05405
%       pchapin@cems.uvm.edu
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%+++++++++++++++++++++++++++++++++
% Preamble and global declarations
%+++++++++++++++++++++++++++++++++
\documentclass[landscape]{slides}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amstext}
\usepackage{stmaryrd}
\usepackage{mathpartir}

\input{slide-macros}
\input{ces_macros}

%++++++++++++++++++++
% The document itself
%++++++++++++++++++++
\begin{document}

%-----------------------
% Title page information
%-----------------------
\title{Risk Assessment in Distributed Authorization\footnote{Based on a paper to appear in \textit{Journal of Computer Security}}}
\author{Peter Chapin, Christian Skalka, X. Sean Wang\\University of Vermont}
\date{March 27, 2007}
\maketitle

%%%%%

%%%%%

\startslide{Outline}

\begin{itemize}

\item Trust Management and $RT_0$
\item $RT^R$ Framework
\item Credential Chain Discovery in $RT^R$
\item Future Work

\end{itemize}

\stopslide

%%%%%

\startslide{Trust Management}
Authorization in a distributed system must be based on general certified attributes, not just identities.
\begin{itemize}
\item Authorizer writes policy describing characteristics of authorized users.
\item Requester provides digitally signed credentials certifying requester's attributes.
\item Authorizer checks if requester has the correct characteristics; that is, \textit{complies with policy}.
\item Authorizer may need to obtain additional credentials dynamically.
\end{itemize}
\stopslide

%%%%%

\startslide{$RT_0$\footnote{Li, Mitchell, Winsborough. \textit{Design of a Role Based Trust Management Framework}, 2002 IEEE Symposium on Security and Privacy}}
Credential forms
\begin{mathpar}
\cred{A.r}{B}{}

\cred{A.r}{B.s}{}

\cred{A.r}{A.s.t}{}

\cred{A.r}{B_1.r_1 \cap B_2.r_2 \cap \cdots \cap B_n.r_n}{}
\end{mathpar}

\begin{itemize}
\item Policies and credentials have the same form.
\item Each principal has a local name space for roles.
\item Meaning of a role, $\mathcal{S}(A.r)$, is the set of entities that are members of that role.
\end{itemize}
\stopslide

%%%%%

\startslide{$RT_0$ Example}
A hotel $H$ wishes to offer discounts to its preferred customers and to members of certain organizations.

\begin{mathpar}
\cred{H.\mathit{discount}}{H.\mathit{preferred}}{}

\cred{H.\mathit{discount}}{H.\mathit{orgs}.\mathit{members}}{}

\cred{H.\mathit{orgs}}{\mathit{AAA}}{}
\end{mathpar}

A later marketing decision by $H$ adds $\cred{H.\mathit{preferred}}{\mathit{AAA}.\mathit{members}}{}$.

Mary has credential $\cred{\mathit{AAA}.\mathit{members}}{M}{}$. This proves compliance with policy two different ways.
\stopslide

%%%%%

\startslide{Example Credential Graph}
\input rt0example
\centerline{\box\graph}

\begin{mathpar}
\cred{H.d}{H.p}{}

\cred{H.d}{H.o.m}{}

\cred{H.p}{A.m}{}

\cred{H.o}{A}{}

\cred{A.m}{M}{}
\end{mathpar}
\stopslide

%%%%%

\startslide{Problem}

Not all credentials are created equal.

\begin{itemize}
\item Some might be signed by questionable keys.
\item Some might be near expiration.
\item Some might be assumed to exist, but not actually be in hand.
\end{itemize}

Existing trust management systems regard credentials as either completely valid or completely invalid. \textit{This is not realistic.}
\stopslide

%%%%%

\startslide{Introducing Risk}

Assigning risks to credentials gives a way to express uncertainties about the credentials.

\begin{itemize}
\item Credentials signed by marginal authorities have high risk.
\item Risk of a credential might increase as its expiration time approaches. For example: $R = R_0(2^{(t - T_i)/(T_e - T_i)} - 1)$
\item Credentials that are presumed to exist or that have unchecked signatures have high risk.
\item Credentials that are part of local policy have very low risk.
\end{itemize}

\stopslide

%%%%%

\startslide{$RT^R$ Framework}

$RT^R$ extends $RT_0$ by assigned risk values to credentials.

\begin{itemize}
\item Let $(\mathcal{K}, \po)$ be a complete lattice over some set $\mathcal{K}$ of risk values with partial ordering $\po$.
\item Credentials now $\cred{A.r}{f}{\risk},\,\risk \in \mathcal{K}$
\item Let $\oplus$ be an associative, commutative, monotonic \textit{risk aggregration operator} over $\mathcal{K}$.
\item Meaning of a role is now a set of risk associations called a \textit{risk assessment}. $\mathcal{S}(A.r) = \{(B, \risk_1), (B, \risk_2), (C, \risk_1)\}$
\end{itemize}

\stopslide

%%%%%

\startslide{Canonical Risk Assessments}

\begin{itemize}
\item Equivalence of risk assessements: $R \cup \{(A, \risk_1), (A, \risk_2)\} = R \cup \{(A, \risk_1)\}$ where $\risk_1 \po \risk_2$.
\item A risk assessment $R$ is \textit{canonical} if there is no $(A, \risk_1), (A, \risk_2) \in R$ such that $\risk_1 \po \risk_2$.
\item Thus any equivalence class of risk assessments has a unique canonical form. \textit{Use this canonical form to represent the meaning of a role.}
\end{itemize}

\stopslide

%%%%%
\startslide{Credential Graph Cycles}
Canonical risk assessments are finite even with cycles in the credential graph.

\input cycle
\centerline{\box\graph}

$\mathcal{S}(A.r) = \{(E, 1), (E, 10)\} = \{(E, 1)\}$
\stopslide
%%%%%

\startslide{Example Revisited}

\input rtrexample
\centerline{\box\graph}

$\mathcal{S}(H.d) = \{(M, 19)\}$

\stopslide

%%%%%

\startslide{Bounded Proof Search}

Given a collection of credentials find a \textit{credential chain} that proves some entity $E$ is in a particular role $A.r$ with a bounded risk.

Abort search in directions where risk is too high.

\begin{itemize}
\item Reduces searching and speeds up the authorization decision.
\item In a distributed search, one may be able to avoid fetching credentials that are not useful.
\item If risks represent wait times, the search finds a credential chain where no certificate takes longer than a given bound to verify.
\end{itemize}

\stopslide

%%%%%

\startslide{Search Algorithm}

Algorithm is a modification of that in [Li et. al.]\footnote{Li, Winsborough, Mitchell, \textit{Distributed Chain Discovery in Trust Management}, Journal of Computer Security, February 2003}

\begin{itemize}
\item Modified breadth-first-search of credential graph.
\item Starts at role $A.r$ and works toward the entities.
\item Graph mutates as search progresses (derived edges added).
\item Accumulated risks tracked; search abandoned where risks excessive.
\end{itemize}

\stopslide

%%%%%

\startslide{Search Algorithm Example: 1}

\input search01
\centerline{\box\graph}

$\risk_M = 20$
\stopslide

%%%%%

\startslide{Search Algorithm Example: 2}

\input search02
\centerline{\box\graph}

$\risk_M = 20$
\stopslide

%%%%%

\startslide{Search Algorithm Example: 3}

\input search03
\centerline{\box\graph}

$\risk_M = 20$
\stopslide

%%%%%
\startslide{Search Algorithm Example: 4}
\input search04
\centerline{\box\graph}

$\risk_M = 20$
\stopslide
%%%%%

\startslide{Search Algorithm Example: 5}

\input search05
\centerline{\box\graph}

$\risk_M = 20$
\stopslide

%%%%%

\startslide{Future Work}

\begin{itemize}
\item Who assigns risks: the authorizer or the issuer?
\item Can risks be used to encode certificate validity concerns?
\item Can risks be used to do cost/benefit analysis?
\begin{itemize}
\item Let risk values have the form $(\risk, t)$
\item Let $(\risk_1, t_1) \po (\risk_2, t_2) \Leftrightarrow (\risk_1 \po \risk_2) \wedge (t_1 \po t_2)$
\item If a search fails, one can try again raising either $\risk$ or $t$ in the threshold.
\end{itemize}
\item How well does $RT^R$ work in the real world?
\end{itemize}

\stopslide

%%%%%

\startslide{Questions?}
\centerline{http://www.cs.uvm.edu/$\sim$pchapin/RMDA.pdf}
\stopslide

\end{document}
